\nwfilename{nw/tmp.nw}\nwbegindocs{0}% -*- mode: poly-noweb; noweb-code-mode: sml-mode; -*-% ===> this file was generated automatically by noweave --- better not edit it

\chapter{Unit Testing}

As we program, it is useful to write down ``companion code'' which gives
example uses of what we're trying to implement. This can be used to
\emph{test} the code we're implementing, to make sure we're ``doing what
we think we're doing''. Before we get to the matter of \emph{how to
write unit tests}, we first need a way to write unit tests (which
programmers call a ``unit-testing framework'').

We begin by implementing a unit-testing framework. This requires having
an {\Tt{}\nwlinkedidentq{assert}{NW4EbhvV-2iEgXD-1}\ :\ string\ ->\ bool\ ->\ unit\nwendquote} function to check if something
has worked as expected (the second parameter will be {\Tt{}true\nwendquote} upon success).
Failure raises a new exception.

We will organize tests into a composite pattern, consisting of
individual assertions wrapped in a {\Tt{}\nwlinkedidentq{TestCase}{NW4EbhvV-4YRS1G-1}\nwendquote} and a collection of
tests bundled together in a {\Tt{}\nwlinkedidentq{TestSuite}{NW4EbhvV-4YRS1G-1}\nwendquote}.

\nwenddocs{}\nwbegincode{1}\sublabel{NW4EbhvV-kKwlF-1}\nwmargintag{{\nwtagstyle{}\subpageref{NW4EbhvV-kKwlF-1}}}\moddef{xunit.sml~{\nwtagstyle{}\subpageref{NW4EbhvV-kKwlF-1}}}\endmoddef\nwstartdeflinemarkup\nwenddeflinemarkup
\LA{}Define \code{}assert\edoc{}~{\nwtagstyle{}\subpageref{NW4EbhvV-2iEgXD-1}}\RA{}
\LA{}Composite test data pattern~{\nwtagstyle{}\subpageref{NW4EbhvV-4YRS1G-1}}\RA{}
\LA{}Define test result data \SMLkwd{type}~{\nwtagstyle{}\subpageref{NW4EbhvV-4L7MKl-1}}\RA{}
\LA{}Run tests and produce results~{\nwtagstyle{}\subpageref{NW4EbhvV-GInkH-1}}\RA{}
\LA{}Print test results to the terminal~{\nwtagstyle{}\subpageref{NW4EbhvV-21Krqt-1}}\RA{}
\LA{}Test suite DSL code~{\nwtagstyle{}\subpageref{NW4EbhvV-1C3sOO-1}}\RA{}
\nwnotused{xunit.sml}\nwendcode{}\nwbegindocs{2}\nwdocspar

We will want to write unit tests, organized into ``folders'' or
\emph{test suites}. Each unit test contains one assertion, which checks
that some condition is true. Some example of simple unit tests of
arithmetic:

\begin{sml}
(* Signatures for the constructors of test suites and unit tests:
  
  suite : string -> Test list -> Test
  test : string -> (unit -> unit) -> Test *)
val ex_tests
= suite "Arithmetic Tests"
        [test "Arithmetic test 1" (fn () => assert "1+1=2" (1+1=2)),
         test "Arithmetic test 2" (fn () => assert "1+1=3" (1+1=3))];
\end{sml}

The {\Tt{}\nwlinkedidentq{assert}{NW4EbhvV-2iEgXD-1}\nwendquote} function takes a label (string) and a Boolean value,
where ``true'' reflects the test does what we expect and ``fail''
reflects\dots well, the test failed. Upon failure, we will raise an
exception for the testing system to catch and interpret as a test failure.

Although we could \emph{reuse} the {\Tt{}Fail\nwendquote} exception, this may also
cause confusion if the code being tested raises a {\Tt{}Fail\nwendquote} exception to
handle particular situations. This leads us to implement our own
{\Tt{}\nwlinkedidentq{AssertionFailure}{NW4EbhvV-2iEgXD-1}\nwendquote} exception. 

\nwenddocs{}\nwbegincode{3}\sublabel{NW4EbhvV-2iEgXD-1}\nwmargintag{{\nwtagstyle{}\subpageref{NW4EbhvV-2iEgXD-1}}}\moddef{Define \code{}assert\edoc{}~{\nwtagstyle{}\subpageref{NW4EbhvV-2iEgXD-1}}}\endmoddef\nwstartdeflinemarkup\nwusesondefline{\\{NW4EbhvV-kKwlF-1}}\nwenddeflinemarkup
\SMLkwd{exception} \nwlinkedidentc{AssertionFailure}{NW4EbhvV-2iEgXD-1} \SMLkwd{of} \SMLtype{string};

\SMLkwd{fun} \nwlinkedidentc{assert}{NW4EbhvV-2iEgXD-1} (msg : \SMLtype{string}) (is_success : \SMLtype{bool}) : \SMLtype{unit} =
    \SMLkwd{if} is_success \SMLkwd{then} () \SMLkwd{else} \SMLkwd{raise} \nwlinkedidentc{AssertionFailure}{NW4EbhvV-2iEgXD-1} msg;
\nwindexdefn{\nwixident{AssertionFailure}}{AssertionFailure}{NW4EbhvV-2iEgXD-1}\nwindexdefn{\nwixident{assert}}{assert}{NW4EbhvV-2iEgXD-1}\eatline
\nwused{\\{NW4EbhvV-kKwlF-1}}\nwidentdefs{\\{{\nwixident{assert}}{assert}}\\{{\nwixident{AssertionFailure}}{AssertionFailure}}}\nwendcode{}\nwbegindocs{4}\nwdocspar
We use a composite design pattern when implementing {\Tt{}\nwlinkedidentq{Test}{NW4EbhvV-4YRS1G-1}\nwendquote}s. 

Now, unit tests are {\Tt{}\nwlinkedidentq{TestCase}{NW4EbhvV-4YRS1G-1}\nwendquote} instances. An individual test case will
take a label (a {\Tt{}string\nwendquote}) and a delayed test procedure ({\Tt{}unit\ ->\ unit\nwendquote}).

A collection of {\Tt{}\nwlinkedidentq{Test}{NW4EbhvV-4YRS1G-1}\nwendquote} instances are collected in a {\Tt{}\nwlinkedidentq{TestSuite}{NW4EbhvV-4YRS1G-1}\nwendquote}.

\nwenddocs{}\nwbegincode{5}\sublabel{NW4EbhvV-4YRS1G-1}\nwmargintag{{\nwtagstyle{}\subpageref{NW4EbhvV-4YRS1G-1}}}\moddef{Composite test data pattern~{\nwtagstyle{}\subpageref{NW4EbhvV-4YRS1G-1}}}\endmoddef\nwstartdeflinemarkup\nwusesondefline{\\{NW4EbhvV-kKwlF-1}}\nwenddeflinemarkup
\SMLkwd{datatype} \nwlinkedidentc{Test}{NW4EbhvV-4YRS1G-1} = \nwlinkedidentc{TestCase}{NW4EbhvV-4YRS1G-1} \SMLkwd{of} \SMLtype{string}*(\SMLtype{unit} -> \SMLtype{unit})
              | \nwlinkedidentc{TestSuite}{NW4EbhvV-4YRS1G-1} \SMLkwd{of} \SMLtype{string}*(\nwlinkedidentc{Test}{NW4EbhvV-4YRS1G-1} list);
\nwindexdefn{\nwixident{Test}}{Test}{NW4EbhvV-4YRS1G-1}\nwindexdefn{\nwixident{TestCase}}{TestCase}{NW4EbhvV-4YRS1G-1}\nwindexdefn{\nwixident{TestSuite}}{TestSuite}{NW4EbhvV-4YRS1G-1}\eatline
\nwused{\\{NW4EbhvV-kKwlF-1}}\nwidentdefs{\\{{\nwixident{Test}}{Test}}\\{{\nwixident{TestCase}}{TestCase}}\\{{\nwixident{TestSuite}}{TestSuite}}}\nwendcode{}\nwbegindocs{6}\nwdocspar
When we run unit tests, we either deal with a single assertion (a
{\Tt{}\nwlinkedidentq{TestCase}{NW4EbhvV-4YRS1G-1}\nwendquote}) or a sequence of tests. For now, we will not print
any information concerning success: only failure or exceptions require
the user's attention. We will give some pseuodocode intuition of what
this would look like:

\begin{sml}
(* run : Test -> unit *)
fun run (TestCase (name, assertion))
    = (assertion()
       handle AssertionFailure msg => print (concat ["Test ",
                                                     name,
                                                     " failed: ",
                                                     msg])
            | e => print (concat ["Unhandled exception in test",
                                  name,
                                  ": ",
                                  exnMessage e]))
  | run (TestSuite (name, tests)) = app run tests;
\end{sml}

We really want to have {\Tt{}run\nwendquote} produce a data structure storing the
results of each test. Towards that end, we need a composite design
pattern for the test results.

An individual test can succeed, fail, or encounter an unhandled
exception.

The {\Tt{}\nwlinkedidentq{TestResult}{NW4EbhvV-4L7MKl-1}\nwendquote} data type is then just a composite design pattern,
whose nodes are just results of individual unit tests. We will store the
amount of time it took to run the unit test (or test suite), the name of
the unit test (or test) suite being executed, and the outcome(s).

\nwenddocs{}\nwbegincode{7}\sublabel{NW4EbhvV-4L7MKl-1}\nwmargintag{{\nwtagstyle{}\subpageref{NW4EbhvV-4L7MKl-1}}}\moddef{Define test result data type~{\nwtagstyle{}\subpageref{NW4EbhvV-4L7MKl-1}}}\endmoddef\nwstartdeflinemarkup\nwusesondefline{\\{NW4EbhvV-kKwlF-1}}\nwenddeflinemarkup
\SMLcomment{(* Extend the framework to \SMLkwd{handle} printing \nwlinkedidentc{test}{NW4EbhvV-4ZJPeu-1} outcomes. *)}
\SMLkwd{datatype} \nwlinkedidentc{TestOutcome}{NW4EbhvV-4L7MKl-1} = TestSuccess
                     | TestFail \SMLkwd{of} \SMLtype{string}
                     | TestException \SMLkwd{of} exn;

\SMLkwd{datatype} \nwlinkedidentc{TestResult}{NW4EbhvV-4L7MKl-1} = \nwlinkedidentc{ResultCase}{NW4EbhvV-4L7MKl-1} \SMLkwd{of} \SMLtype{string}*Time.time*\nwlinkedidentc{TestOutcome}{NW4EbhvV-4L7MKl-1}
                    | \nwlinkedidentc{ResultSuite}{NW4EbhvV-4L7MKl-1} \SMLkwd{of} \SMLtype{string}*Time.time*(\nwlinkedidentc{TestResult}{NW4EbhvV-4L7MKl-1} list);
\nwindexdefn{\nwixident{TestOutcome}}{TestOutcome}{NW4EbhvV-4L7MKl-1}\nwindexdefn{\nwixident{TestResult}}{TestResult}{NW4EbhvV-4L7MKl-1}\nwindexdefn{\nwixident{ResultCase}}{ResultCase}{NW4EbhvV-4L7MKl-1}\nwindexdefn{\nwixident{ResultSuite}}{ResultSuite}{NW4EbhvV-4L7MKl-1}\eatline
\nwused{\\{NW4EbhvV-kKwlF-1}}\nwidentdefs{\\{{\nwixident{ResultCase}}{ResultCase}}\\{{\nwixident{ResultSuite}}{ResultSuite}}\\{{\nwixident{TestOutcome}}{TestOutcome}}\\{{\nwixident{TestResult}}{TestResult}}}\nwidentuses{\\{{\nwixident{test}}{test}}}\nwindexuse{\nwixident{test}}{test}{NW4EbhvV-4L7MKl-1}\nwendcode{}\nwbegindocs{8}\nwdocspar
Running unit tests will then produce a {\Tt{}\nwlinkedidentq{TestResult}{NW4EbhvV-4L7MKl-1}\nwendquote} instance storing
the germane information.

\nwenddocs{}\nwbegincode{9}\sublabel{NW4EbhvV-GInkH-1}\nwmargintag{{\nwtagstyle{}\subpageref{NW4EbhvV-GInkH-1}}}\moddef{Run tests and produce results~{\nwtagstyle{}\subpageref{NW4EbhvV-GInkH-1}}}\endmoddef\nwstartdeflinemarkup\nwusesondefline{\\{NW4EbhvV-kKwlF-1}}\nwenddeflinemarkup
\SMLcomment{(* \nwlinkedidentc{run_test}{NW4EbhvV-GInkH-1} : \nwlinkedidentc{Test}{NW4EbhvV-4YRS1G-1} -> \nwlinkedidentc{TestResult}{NW4EbhvV-4L7MKl-1} *)}
\SMLkwd{fun} \nwlinkedidentc{run_test}{NW4EbhvV-GInkH-1} (\nwlinkedidentc{test}{NW4EbhvV-4ZJPeu-1} : \nwlinkedidentc{Test}{NW4EbhvV-4YRS1G-1}) : \nwlinkedidentc{TestResult}{NW4EbhvV-4L7MKl-1} =
\SMLkwd{let}
  \SMLkwd{val} start = Time.now()
\SMLkwd{in}
  \SMLkwd{case} \nwlinkedidentc{test}{NW4EbhvV-4ZJPeu-1} \SMLkwd{of}
      \LA{}Run an individual \SMLtype{unit} test~{\nwtagstyle{}\subpageref{NW4EbhvV-3DPdJC-1}}\RA{}
    | \LA{}Run a test suite~{\nwtagstyle{}\subpageref{NW4EbhvV-47ZbGR-1}}\RA{}
\SMLkwd{end};
\nwindexdefn{\nwixident{run{\_}test}}{run:untest}{NW4EbhvV-GInkH-1}\eatline
\nwused{\\{NW4EbhvV-kKwlF-1}}\nwidentdefs{\\{{\nwixident{run{\_}test}}{run:untest}}}\nwidentuses{\\{{\nwixident{Test}}{Test}}\\{{\nwixident{test}}{test}}\\{{\nwixident{TestResult}}{TestResult}}}\nwindexuse{\nwixident{Test}}{Test}{NW4EbhvV-GInkH-1}\nwindexuse{\nwixident{test}}{test}{NW4EbhvV-GInkH-1}\nwindexuse{\nwixident{TestResult}}{TestResult}{NW4EbhvV-GInkH-1}\nwendcode{}\nwbegindocs{10}\nwdocspar
If the assertion runs as expected, we record a {\Tt{}TestSuccess\nwendquote} for the
{\Tt{}case{\_}name\nwendquote}, along with how long it took.
However, if an {\Tt{}\nwlinkedidentq{AssertionFailure}{NW4EbhvV-2iEgXD-1}\nwendquote} is raised, then the test failed, and
we record a failure for the {\Tt{}case{\_}name\nwendquote}.
Worse, if any other exception occurs, we should record something
exceptional has happened (and store the details of the exception raised).

\nwenddocs{}\nwbegincode{11}\sublabel{NW4EbhvV-3DPdJC-1}\nwmargintag{{\nwtagstyle{}\subpageref{NW4EbhvV-3DPdJC-1}}}\moddef{Run an individual unit test~{\nwtagstyle{}\subpageref{NW4EbhvV-3DPdJC-1}}}\endmoddef\nwstartdeflinemarkup\nwusesondefline{\\{NW4EbhvV-GInkH-1}}\nwenddeflinemarkup
(\nwlinkedidentc{TestCase}{NW4EbhvV-4YRS1G-1} (case_name, assertion)) =>
((assertion();
  \nwlinkedidentc{ResultCase}{NW4EbhvV-4L7MKl-1} (case_name, (Time.-)(Time.now(), start), TestSuccess))
 \SMLkwd{handle} \nwlinkedidentc{AssertionFailure}{NW4EbhvV-2iEgXD-1} msg => \nwlinkedidentc{ResultCase}{NW4EbhvV-4L7MKl-1} (case_name,
                                            (Time.-)(Time.now(), start),
                                            TestFail msg)
      | e => \nwlinkedidentc{ResultCase}{NW4EbhvV-4L7MKl-1} (case_name,
                         (Time.-)(Time.now(), start),
                         TestException e))
\nwused{\\{NW4EbhvV-GInkH-1}}\nwidentuses{\\{{\nwixident{AssertionFailure}}{AssertionFailure}}\\{{\nwixident{ResultCase}}{ResultCase}}\\{{\nwixident{TestCase}}{TestCase}}}\nwindexuse{\nwixident{AssertionFailure}}{AssertionFailure}{NW4EbhvV-3DPdJC-1}\nwindexuse{\nwixident{ResultCase}}{ResultCase}{NW4EbhvV-3DPdJC-1}\nwindexuse{\nwixident{TestCase}}{TestCase}{NW4EbhvV-3DPdJC-1}\nwendcode{}\nwbegindocs{12}\nwdocspar

Running all the tests on a test suite amounts to iterating through each
element in the suite, and running each element recursively. We use the
fact that Standard ML is call-by-value to have {\Tt{}results\nwendquote} be computed
first, and upon completion the elapsed time interval {\Tt{}dt\nwendquote} is determined.
Thus a test suite's results will include the time it took to run every
``descendent'' test contained within it.

\nwenddocs{}\nwbegincode{13}\sublabel{NW4EbhvV-47ZbGR-1}\nwmargintag{{\nwtagstyle{}\subpageref{NW4EbhvV-47ZbGR-1}}}\moddef{Run a test suite~{\nwtagstyle{}\subpageref{NW4EbhvV-47ZbGR-1}}}\endmoddef\nwstartdeflinemarkup\nwusesondefline{\\{NW4EbhvV-GInkH-1}}\nwenddeflinemarkup
(\nwlinkedidentc{TestSuite}{NW4EbhvV-4YRS1G-1} (suite_name, tests)) =>
(\SMLkwd{let}
    \SMLkwd{val} results = map \nwlinkedidentc{run_test}{NW4EbhvV-GInkH-1} tests
    \SMLkwd{val} dt = (Time.-)(Time.now(), start)
\SMLkwd{in}
    \nwlinkedidentc{ResultSuite}{NW4EbhvV-4L7MKl-1} (suite_name, dt, results)
\SMLkwd{end})
\nwused{\\{NW4EbhvV-GInkH-1}}\nwidentuses{\\{{\nwixident{ResultSuite}}{ResultSuite}}\\{{\nwixident{run{\_}test}}{run:untest}}\\{{\nwixident{TestSuite}}{TestSuite}}}\nwindexuse{\nwixident{ResultSuite}}{ResultSuite}{NW4EbhvV-47ZbGR-1}\nwindexuse{\nwixident{run{\_}test}}{run:untest}{NW4EbhvV-47ZbGR-1}\nwindexuse{\nwixident{TestSuite}}{TestSuite}{NW4EbhvV-47ZbGR-1}\nwendcode{}\nwbegindocs{14}\nwdocspar

We can then implement a Reporter, which will report and/or record
information concerning the test results. For our purposes, we only need
to print out to the screen the test results. Again, since we're
processing a composite data pattern, we can implement this as two cases:
one for the leafs in the collection, another for the test suites.

Later it may be useful to store this information in a separate file
(Java routinely uses XML files to store results from running unit
tests).
There are other possibilities, but for now we will just print the tests
to the terminal.

It has become conventional to report time intervals to microseconds. We
do this with the following function. The {\Tt{}Time.fmt\nwendquote} will format the
time interval to {\Tt{}n\nwendquote} digits after the decimal point, in units of seconds.

\nwenddocs{}\nwbegincode{15}\sublabel{NW4EbhvV-21Krqt-1}\nwmargintag{{\nwtagstyle{}\subpageref{NW4EbhvV-21Krqt-1}}}\moddef{Print test results to the terminal~{\nwtagstyle{}\subpageref{NW4EbhvV-21Krqt-1}}}\endmoddef\nwstartdeflinemarkup\nwusesondefline{\\{NW4EbhvV-kKwlF-1}}\nwenddeflinemarkup
\SMLcomment{(* Print the results to the terminal.

    \nwlinkedidentc{report_results}{NW4EbhvV-21Krqt-1} : \nwlinkedidentc{TestResult}{NW4EbhvV-4L7MKl-1} -> \SMLtype{unit} *)}
\SMLkwd{fun} \nwlinkedidentc{report_results}{NW4EbhvV-21Krqt-1} (\nwlinkedidentc{ResultCase}{NW4EbhvV-4L7MKl-1} (name,dt,outcome)) =
    \LA{}Print the result for a test \SMLkwd{case} to the terminal~{\nwtagstyle{}\subpageref{NW4EbhvV-EoKaK-1}}\RA{}
  | \nwlinkedidentc{report_results}{NW4EbhvV-21Krqt-1} (\nwlinkedidentc{ResultSuite}{NW4EbhvV-4L7MKl-1} (name,dt,outcomes)) =
    (print (concat ["Suite ", name, " (", Time.fmt 3 dt, ")\\n"]);
     app \nwlinkedidentc{report_results}{NW4EbhvV-21Krqt-1} outcomes);
\nwindexdefn{\nwixident{report{\_}results}}{report:unresults}{NW4EbhvV-21Krqt-1}\eatline
\nwused{\\{NW4EbhvV-kKwlF-1}}\nwidentdefs{\\{{\nwixident{report{\_}results}}{report:unresults}}}\nwidentuses{\\{{\nwixident{ResultCase}}{ResultCase}}\\{{\nwixident{ResultSuite}}{ResultSuite}}\\{{\nwixident{TestResult}}{TestResult}}}\nwindexuse{\nwixident{ResultCase}}{ResultCase}{NW4EbhvV-21Krqt-1}\nwindexuse{\nwixident{ResultSuite}}{ResultSuite}{NW4EbhvV-21Krqt-1}\nwindexuse{\nwixident{TestResult}}{TestResult}{NW4EbhvV-21Krqt-1}\nwendcode{}\nwbegindocs{16}\nwdocspar
The outcome for an individual unit test is a success, failure, or an exception.
Usually, we only care about failures, but it is good to know that a test
has been executed.

\nwenddocs{}\nwbegincode{17}\sublabel{NW4EbhvV-EoKaK-1}\nwmargintag{{\nwtagstyle{}\subpageref{NW4EbhvV-EoKaK-1}}}\moddef{Print the result for a test case to the terminal~{\nwtagstyle{}\subpageref{NW4EbhvV-EoKaK-1}}}\endmoddef\nwstartdeflinemarkup\nwusesondefline{\\{NW4EbhvV-21Krqt-1}}\nwenddeflinemarkup
(\SMLkwd{case} outcome \SMLkwd{of}
     TestSuccess => print (concat ["\nwlinkedidentc{Test}{NW4EbhvV-4YRS1G-1} ",
                                   name,
                                   " (",
                                   Time.fmt 3 dt,
                                   ")",
                                   ": SUCCESS\\n"])
   | (TestFail msg) => print (concat ["\nwlinkedidentc{Test}{NW4EbhvV-4YRS1G-1} ",
                                      name,
                                      " (",
                                      Time.fmt 3 dt,
                                      "): FAIL -- ",
                                      msg,
                                      "\\n"])
   | (TestException e) => print (concat ["\nwlinkedidentc{Test}{NW4EbhvV-4YRS1G-1} ",
                                         name,
                                         " (",
                                         Time.fmt 3 dt,
                                         "): UNHANDLED EXCEPTION ",
                                         exnMessage e,
                                         "\\n"]))
\nwused{\\{NW4EbhvV-21Krqt-1}}\nwidentuses{\\{{\nwixident{Test}}{Test}}}\nwindexuse{\nwixident{Test}}{Test}{NW4EbhvV-EoKaK-1}\nwendcode{}\nwbegindocs{18}\nwdocspar


\section{Test and Suite Functions}

We can now introduce helper functions to make implementing unit tests
easier. The programmer is intended to write unit tests in separate
files, each file containing exactly one suite.

The basic idea is that the programmer will just define the current
suite, then write down unit tests to be accumulated in the suite. For
example:
\begin{sml}
suite "FooBarTests";

test "BaazTest1" fn () => (* ... *);

test "BaazTest2" fn () => (* ... *);

(* etc. *)
\end{sml}
Towards that end, we will track the tests being defined in the current
suite, and append the suite in the collection of all suites.

Ideally, this would be done using some mutable dictionary construction,
but we will just use a list. There might be two suites sharing the same
name (which should be collapsed and merged together), but we will just
accept duplicate suites as a sad fact of life rather than a bug that
must be fixed.

\nwenddocs{}\nwbegincode{19}\sublabel{NW4EbhvV-1C3sOO-1}\nwmargintag{{\nwtagstyle{}\subpageref{NW4EbhvV-1C3sOO-1}}}\moddef{Test suite DSL code~{\nwtagstyle{}\subpageref{NW4EbhvV-1C3sOO-1}}}\endmoddef\nwstartdeflinemarkup\nwusesondefline{\\{NW4EbhvV-kKwlF-1}}\nwenddeflinemarkup
\SMLkwd{val} \nwlinkedidentc{current_suite}{NW4EbhvV-1C3sOO-1} : (\SMLtype{string}*(\nwlinkedidentc{Test}{NW4EbhvV-4YRS1G-1} list)) ref = ref ("", []);
\SMLkwd{val} \nwlinkedidentc{all_suites}{NW4EbhvV-1C3sOO-1} : (\nwlinkedidentc{Test}{NW4EbhvV-4YRS1G-1} list) ref = ref [];

\LA{}Append the current suite to all test suites~{\nwtagstyle{}\subpageref{NW4EbhvV-CI46F-1}}\RA{}
  
\LA{}Start a new test suite~{\nwtagstyle{}\subpageref{NW4EbhvV-q8fSD-1}}\RA{}
  
\LA{}Construct a new test \SMLkwd{case}~{\nwtagstyle{}\subpageref{NW4EbhvV-4ZJPeu-1}}\RA{}
\nwindexdefn{\nwixident{current{\_}suite}}{current:unsuite}{NW4EbhvV-1C3sOO-1}\nwindexdefn{\nwixident{all{\_}suites}}{all:unsuites}{NW4EbhvV-1C3sOO-1}\eatline
\nwused{\\{NW4EbhvV-kKwlF-1}}\nwidentdefs{\\{{\nwixident{all{\_}suites}}{all:unsuites}}\\{{\nwixident{current{\_}suite}}{current:unsuite}}}\nwidentuses{\\{{\nwixident{Test}}{Test}}}\nwindexuse{\nwixident{Test}}{Test}{NW4EbhvV-1C3sOO-1}\nwendcode{}\nwbegindocs{20}\nwdocspar
We will prepend the current suite to the collection of all test
suites. This will also initialize an empty list of tests and an
empty-named suite.

\nwenddocs{}\nwbegincode{21}\sublabel{NW4EbhvV-CI46F-1}\nwmargintag{{\nwtagstyle{}\subpageref{NW4EbhvV-CI46F-1}}}\moddef{Append the current suite to all test suites~{\nwtagstyle{}\subpageref{NW4EbhvV-CI46F-1}}}\endmoddef\nwstartdeflinemarkup\nwusesondefline{\\{NW4EbhvV-1C3sOO-1}}\nwenddeflinemarkup
\SMLkwd{fun} \nwlinkedidentc{append_suite}{NW4EbhvV-CI46F-1} \SMLtype{unit} =
    \SMLkwd{let}
        \SMLkwd{val} (name, tests) = !\nwlinkedidentc{current_suite}{NW4EbhvV-1C3sOO-1}
    \SMLkwd{in}
        \nwlinkedidentc{all_suites}{NW4EbhvV-1C3sOO-1} := (\nwlinkedidentc{TestSuite}{NW4EbhvV-4YRS1G-1} (name, tests)) :: (!\nwlinkedidentc{all_suites}{NW4EbhvV-1C3sOO-1});
        \nwlinkedidentc{current_suite}{NW4EbhvV-1C3sOO-1} := ("", [])
    \SMLkwd{end};
\nwindexdefn{\nwixident{append{\_}suite}}{append:unsuite}{NW4EbhvV-CI46F-1}\eatline
\nwused{\\{NW4EbhvV-1C3sOO-1}}\nwidentdefs{\\{{\nwixident{append{\_}suite}}{append:unsuite}}}\nwidentuses{\\{{\nwixident{all{\_}suites}}{all:unsuites}}\\{{\nwixident{current{\_}suite}}{current:unsuite}}\\{{\nwixident{TestSuite}}{TestSuite}}}\nwindexuse{\nwixident{all{\_}suites}}{all:unsuites}{NW4EbhvV-CI46F-1}\nwindexuse{\nwixident{current{\_}suite}}{current:unsuite}{NW4EbhvV-CI46F-1}\nwindexuse{\nwixident{TestSuite}}{TestSuite}{NW4EbhvV-CI46F-1}\nwendcode{}\nwbegindocs{22}\nwdocspar
Then the domain-specific code for starting a test suite just amounts to
calling {\Tt{}\nwlinkedidentq{append{\_}suite}{NW4EbhvV-CI46F-1}\nwendquote}, but updating the {\Tt{}\nwlinkedidentq{current{\_}suite}{NW4EbhvV-1C3sOO-1}\nwendquote} to have the
given {\Tt{}name\nwendquote} argument.

\nwenddocs{}\nwbegincode{23}\sublabel{NW4EbhvV-q8fSD-1}\nwmargintag{{\nwtagstyle{}\subpageref{NW4EbhvV-q8fSD-1}}}\moddef{Start a new test suite~{\nwtagstyle{}\subpageref{NW4EbhvV-q8fSD-1}}}\endmoddef\nwstartdeflinemarkup\nwusesondefline{\\{NW4EbhvV-1C3sOO-1}}\nwenddeflinemarkup
\SMLkwd{fun} \nwlinkedidentc{suite}{NW4EbhvV-q8fSD-1} (name : \SMLtype{string}) = (
    \nwlinkedidentc{append_suite}{NW4EbhvV-CI46F-1}(name);
    \nwlinkedidentc{current_suite}{NW4EbhvV-1C3sOO-1} := (name, []);
);
\nwindexdefn{\nwixident{suite}}{suite}{NW4EbhvV-q8fSD-1}\eatline
\nwused{\\{NW4EbhvV-1C3sOO-1}}\nwidentdefs{\\{{\nwixident{suite}}{suite}}}\nwidentuses{\\{{\nwixident{append{\_}suite}}{append:unsuite}}\\{{\nwixident{current{\_}suite}}{current:unsuite}}}\nwindexuse{\nwixident{append{\_}suite}}{append:unsuite}{NW4EbhvV-q8fSD-1}\nwindexuse{\nwixident{current{\_}suite}}{current:unsuite}{NW4EbhvV-q8fSD-1}\nwendcode{}\nwbegindocs{24}\nwdocspar
Defining a test case will be domain-specific code to add the test to the
current test suite.

\nwenddocs{}\nwbegincode{25}\sublabel{NW4EbhvV-4ZJPeu-1}\nwmargintag{{\nwtagstyle{}\subpageref{NW4EbhvV-4ZJPeu-1}}}\moddef{Construct a new test case~{\nwtagstyle{}\subpageref{NW4EbhvV-4ZJPeu-1}}}\endmoddef\nwstartdeflinemarkup\nwusesondefline{\\{NW4EbhvV-1C3sOO-1}}\nwenddeflinemarkup
\SMLkwd{fun} \nwlinkedidentc{test}{NW4EbhvV-4ZJPeu-1} (name : \SMLtype{string}) (assertion : \SMLtype{unit} -> \SMLtype{unit}) : \SMLtype{unit} =
    \SMLkwd{let}
        \SMLkwd{val} (suite_name, tests) = !\nwlinkedidentc{current_suite}{NW4EbhvV-1C3sOO-1}
    \SMLkwd{in}
        \nwlinkedidentc{current_suite}{NW4EbhvV-1C3sOO-1} := (suite_name, (\nwlinkedidentc{TestCase}{NW4EbhvV-4YRS1G-1} (name, assertion))::tests)
    \SMLkwd{end};
\nwindexdefn{\nwixident{test}}{test}{NW4EbhvV-4ZJPeu-1}\eatline

\nwused{\\{NW4EbhvV-1C3sOO-1}}\nwidentdefs{\\{{\nwixident{test}}{test}}}\nwidentuses{\\{{\nwixident{current{\_}suite}}{current:unsuite}}\\{{\nwixident{TestCase}}{TestCase}}}\nwindexuse{\nwixident{current{\_}suite}}{current:unsuite}{NW4EbhvV-4ZJPeu-1}\nwindexuse{\nwixident{TestCase}}{TestCase}{NW4EbhvV-4ZJPeu-1}\nwendcode{}

\nwixlogsorted{c}{{Append the current suite to all test suites}{NW4EbhvV-CI46F-1}{\nwixu{NW4EbhvV-1C3sOO-1}\nwixd{NW4EbhvV-CI46F-1}}}%
\nwixlogsorted{c}{{Composite test data pattern}{NW4EbhvV-4YRS1G-1}{\nwixu{NW4EbhvV-kKwlF-1}\nwixd{NW4EbhvV-4YRS1G-1}}}%
\nwixlogsorted{c}{{Construct a new test case}{NW4EbhvV-4ZJPeu-1}{\nwixu{NW4EbhvV-1C3sOO-1}\nwixd{NW4EbhvV-4ZJPeu-1}}}%
\nwixlogsorted{c}{{Define \code{}assert\edoc{}}{NW4EbhvV-2iEgXD-1}{\nwixu{NW4EbhvV-kKwlF-1}\nwixd{NW4EbhvV-2iEgXD-1}}}%
\nwixlogsorted{c}{{Define test result data type}{NW4EbhvV-4L7MKl-1}{\nwixu{NW4EbhvV-kKwlF-1}\nwixd{NW4EbhvV-4L7MKl-1}}}%
\nwixlogsorted{c}{{Print test results to the terminal}{NW4EbhvV-21Krqt-1}{\nwixu{NW4EbhvV-kKwlF-1}\nwixd{NW4EbhvV-21Krqt-1}}}%
\nwixlogsorted{c}{{Print the result for a test case to the terminal}{NW4EbhvV-EoKaK-1}{\nwixu{NW4EbhvV-21Krqt-1}\nwixd{NW4EbhvV-EoKaK-1}}}%
\nwixlogsorted{c}{{Run a test suite}{NW4EbhvV-47ZbGR-1}{\nwixu{NW4EbhvV-GInkH-1}\nwixd{NW4EbhvV-47ZbGR-1}}}%
\nwixlogsorted{c}{{Run an individual unit test}{NW4EbhvV-3DPdJC-1}{\nwixu{NW4EbhvV-GInkH-1}\nwixd{NW4EbhvV-3DPdJC-1}}}%
\nwixlogsorted{c}{{Run tests and produce results}{NW4EbhvV-GInkH-1}{\nwixu{NW4EbhvV-kKwlF-1}\nwixd{NW4EbhvV-GInkH-1}}}%
\nwixlogsorted{c}{{Start a new test suite}{NW4EbhvV-q8fSD-1}{\nwixu{NW4EbhvV-1C3sOO-1}\nwixd{NW4EbhvV-q8fSD-1}}}%
\nwixlogsorted{c}{{Test suite DSL code}{NW4EbhvV-1C3sOO-1}{\nwixu{NW4EbhvV-kKwlF-1}\nwixd{NW4EbhvV-1C3sOO-1}}}%
\nwixlogsorted{c}{{xunit.sml}{NW4EbhvV-kKwlF-1}{\nwixd{NW4EbhvV-kKwlF-1}}}%
\nwixlogsorted{i}{{\nwixident{all{\_}suites}}{all:unsuites}}%
\nwixlogsorted{i}{{\nwixident{append{\_}suite}}{append:unsuite}}%
\nwixlogsorted{i}{{\nwixident{assert}}{assert}}%
\nwixlogsorted{i}{{\nwixident{AssertionFailure}}{AssertionFailure}}%
\nwixlogsorted{i}{{\nwixident{current{\_}suite}}{current:unsuite}}%
\nwixlogsorted{i}{{\nwixident{report{\_}results}}{report:unresults}}%
\nwixlogsorted{i}{{\nwixident{ResultCase}}{ResultCase}}%
\nwixlogsorted{i}{{\nwixident{ResultSuite}}{ResultSuite}}%
\nwixlogsorted{i}{{\nwixident{run{\_}test}}{run:untest}}%
\nwixlogsorted{i}{{\nwixident{suite}}{suite}}%
\nwixlogsorted{i}{{\nwixident{Test}}{Test}}%
\nwixlogsorted{i}{{\nwixident{test}}{test}}%
\nwixlogsorted{i}{{\nwixident{TestCase}}{TestCase}}%
\nwixlogsorted{i}{{\nwixident{TestOutcome}}{TestOutcome}}%
\nwixlogsorted{i}{{\nwixident{TestResult}}{TestResult}}%
\nwixlogsorted{i}{{\nwixident{TestSuite}}{TestSuite}}%

